Databases are an effective and efficient method for storage and access of mass-spectrometry data

### 2025-01-31 manuscript/manuscript.Rmd

```{r}
download.file("https://ftp.ebi.ac.uk/pub/databases/metabolights/studies/public/MTBLS10096/FILES/DERIVED_FILES/POS/Feuille_Extrait_1_ESI_pos_01_7273.mzXML", mode = "wb", destfile = "../massive_data/Feuille_Extrait_1_ESI_pos_01_7273.mzXML")

download.file("ftp://massive.ucsd.edu/v01/MSV000080030/peak/Forensic_study_80_volunteers/Forensic_Hands_mzXML/Forensic_Hands_plate_1_mzXML/Sample/P1_1D10v7_Palm_GD10_01_24008.mzXML", "../massive_data/P1_1D10v7_Palm_GD10_01_24008.mzXML", mode = "wb")

download.file("https://ftp.ebi.ac.uk/pub/databases/metabolights/studies/public/MTBLS417/FILES/CS52684_pos_SWATH.mzXML", mode = "wb", destfile = "../massive_data/CS52684_pos_SWATH.mzXML")
```

@@

```{shell}
cd Documents\Will\mzsql
conda activate mzsql

@@

```{r}
git clone https://github.com/wkumler/mzsql
```

@@

--

@@

Which file types we were able to find/write parsers for

@@

>>>>

@@

```{shell}
cd Documents\Will
git clone https://github.com/wkumler/mzsql
cd mzsql

conda create -n "mzsql_3_11" python=3.11 numpy=1.26.4
pip install https://pypi.cs.uni-tuebingen.de/packages/pyopenms-3.0.0.dev20230306-cp311-cp311-win_amd64.whl#md5=cdce650718bea07a0193b3227a2c075b
conda activate mzsql_3_11
pip install -e .
pip install git+https://github.com/PNNL-m-q/mzapy.git@no_full_mz_array#egg=mzapy

cd java_things
curl -O https://dlcdn.apache.org/maven/maven-3/3.9.9/binaries/apache-maven-3.9.9-bin.zip
tar -xf apache-maven-3.9.9-bin.zip
git clone https://github.com/optimusmoose/MZTree.git
cd MZTree
cd msDataServer
C:\Users\willi\Documents\mzsql\java_things\apache-maven-3.9.9\bin\mvn package
java -Dsun.java2d.d3d=false -Dsun.java2d.opengl=false -jar target\msDataServer-1.0.jar
(manually) open mztree file in port 4568

git clone https://github.com/yrm9837/mzMD-java.git
cd mzMD-java
C:\Users\willi\Documents\mzsql\java_things\apache-maven-3.9.9\bin\mvn package
java -Dsun.java2d.d3d=false -Dsun.java2d.opengl=false -jar target\msDataServer-1.0.jar
(manually) open mzmd file in port 4567
cd ..

@@

@@

$sample_metadata

@@

## Building parsers

Which file types we were able to find/write parsers for and our notes on their methodology

Good documentation for pyteomics, pyopenms, pymzml, mzmlb (at least via pyteomics)
Middling documentation for mza/mzapy, mzDB
Terrible documentation for Aird, MzTree, mzMD, mz5
Nonexistent documentation for YAFMS, Shaduf

pyopenms requires an old version of Python (3.11) and numpy (1.26.4) because of AVX support - discovered via GitHub issue with no requirements.txt file or environment.yml. Therefore requires using old version of the package.

Aird has zero documentation or application support (Python and C# versions are functionally deprecated)

Toffee only supports TOF data

MzTree and mzMD have no MS2+ support and don't support scan extraction
Neither of them handle multiple files

mz5 doesn't make it clear how to access the precursor masses for MS2 things so those parsers are missing, ref mzMLb paper here

mzDB handles metadata as XML strings and doesn't support bounding boxes for MS2 things afaics
Doesn't support .d files (only .raw and .wiff)
Doesn't handle polarity switching data

MZA can't perform centroiding simultaneously with .raw conversion, requires switch from .mzML
MZA doesn't support .wiff (only .raw or .d)
mzapy officially only supports .d (.raw requires installing from a separate development branch)
mzapy doesn't support direct scan extraction - baffling given how well the file type exposes this
mzapy doesn't index by MS2 things (only RT/DT)

Several fundamental inefficiencies identified:
1. Looping over each scan within a file
2. Looping over each file within a dataset
3. Decoding mz/int information each time if filtering on these fields
4. RTs are not accessible outside of looping over each scan for RT subsetting
5. Scan ids are not accessible outside of looping over each scan for scan extraction

## Timings for a single file

## Timings for a dataset with multiple files (and file sizes?)

## Database optimization 
1) Indices and ordering
2) Peak data extraction via joins vs loops vs unified query

@@

Not a single parser was without flaws - all were either slow, poorly documented, or had highly limited functionality

Querying things by scan number is dumb.
1) Different indices for pyopenms vs pymzml vs pyteomics
2) idx = scan_num only works for consecutive scans (no multi-experiment, no polarity switching, no filtering)
3) Only USEFUL when you're parsing the entire file! Nobody knows what scan number 976 corresponds to, making extraction by number useless
